Synthesizing Symmetric Lenses

### 2018-07-02 papers/SymmetricLenses/main.tex

Lens languages, like any domain-specific language, also cut down the space of
possible programs one can write, and the smaller the space of programs,
the easier it is to find the program one is looking for.  
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician} has exploited this fact by demonstrating that it
is possible to synthesize relatively large, complex lenses from specifications.
In particular, given types that describe source and target formats, there are
synthesis algorithms that will find bijective lenses~\cite{optician} for files
in ad-hoc formats, such as those housing linux application configurations.
Oftentimes, these algorithms find such lenses in less than a second, and without
the aid of any examples at all, with just types describing the source and target
formats. \dpw{I would cite the ICFP paper (cite string:
  maina+:quotient-synthesis) as to appear but I'm not sure how to do that
  without possibly violating the spirit of double-blind reviewing. ie, if it
  hasn't appeared yet, how could we know the content if we aren't the authors?
  Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
  2018 website lists the paper; I think it is okay to cite it. The reviewer can
  assume we contacted the authors for a copy. We could ask Stephanie for a
  ruling if we want to be sure. One of the principles of double blind as adopted
  by SIGPLAN is not to reduce the quality of the paper; omitting this citation
  would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
  third person, even if that's a bit awkward, but otherwise citing it is OK.}

@@

The second difficulty with synthesizing symmetric lenses is more
fundamental\dpw{better word?}: While the number of bijective lenses between two
related formats is often tiny, the number of symmetric lenses is often large. In
practice, the difference in cardinality means that when a search algorithm
selects the first bijective lens it finds, that lens has a good chance of being
the one a user wants. In contrast, and to our surprise, our experiments show
that if our search algorithm selects the first symmetric lens it finds, it
typically fails to find the lens the user wants. The flexibity these lenses
admit when it comes to eliding information generates an explosion of possible
transformations. To disambiguate between the many possible lenses, we need new
principles for identifying the ``likely'' ones, and we need new algorithms to
search the space of program transformations using these principles. \dpw{Were
  the bijective lenses chosen using a ``smaller'' criteria? I forget... if not
  edit the prior paragraph.}\afm{no, chosen using a ``first'' criteria.
  Edited.}\dpw{Ideally, we could back up the claim that there are many symmetric
  lenses and not many bijective lenses by graphing the number of such lenses
  somehow. eg, in a past synthesis paper with Steve (POPL 2016) we graphed the
  untyped terms of type $\mathsf{int}\rightarrow\mathsf{int}$. I wonder if we
  can do something similar here.}\afm{What about lenses between ``a''* and
  ''b``*}

To overcome this second difficulty, we turned to information theory to try to
find a way to characterize the likely and unlikely lenses. Intuitively, a
``likely'' lens is one that propagates a lot of information from source to
target and also from target back to source. Conversely, given the source, the
target is ``unlikely'' to require very much additional information in order to
reconstruct it---if it were\bcp{were what? likely, or unlikely?}, then one is
unlikely to be attempting to synthesize a lens between the two formats\bcp{why?
  even if two big formats only share a little bit of common information,
  mightn't one want a lens between them?}! In order to implement this intuition,
we turn to \emph{stochastic regular expressions} (SREs). Stochastic regular
expressions define a set of strings, as do ordinary regular expressions, but
they also provide the expected distribution of strings within that set. Such
distributions allow us to compute the amount of information gained when
presented an element of the set. For example, the language of the SRE
$\lstinline{"a"} |_{0.5}
\lstinline{"b"}$ contains two elements (the strings
\lstinline{"a"} and
\lstinline{"b"}). The probability 0.5 annotating the union indicates that both
elements are equally likely to appear. Consequently, if we observe an element of
that language (either \lstinline{"a"} or \lstinline{"b"}), we will gain one bit
of information. Said another way, one bit, together with the SRE, suffices to
recover the string in question.\bcp{Is this implying that, if the annotation
  were different, we'd need something other than one bit to recover the
  string??}
\afm{From an information theoretic perspective, yes, because that perspective
  views it as the amount of information if we were communicating a stream of
  these choices.} On the other hand, consider the language of the SRE \lstinline{c}. It contains
just one element (the string $c$). If we observe that our algorithm generates
$\lstinline{"c"}$, we gain zero bits of information --- the string was already determined.
Hence\bcp{?}, a lens is ``likely,'' in a technical sense, if relative to the
target SRE $T$, very little information from $t$ is required to generate it.

@@

In summary, our main contributions are:\bcp{annotate with section numbers}
\begin{itemize}
\item We define the space of simple symmetric lenses, which lies between
  asymmetric lenses and full symmetric lenses and has the useful property that
  they don't have any internal state (\S\ref{sec:overview}). We characterize the
  simple lenses semantically as the subset of full symmetric lenses for which a
  ``forgetfulness'' property holds (\S\ref{sec:related}). We also analyze the
  gap between simple and full symmetric lenses empirically and argue that simple
  ones suffice for many applications (\S\ref{sec:evaluation}).
\item We show how to synthesize simple symmetric lenses (including all asymmetric lenses as a special case),
using a novel application of stochastic regular expressions to guide the search process.  In particular, we
show that the RE star semi-ring equivalences remain valid when their semantics
are enriched by distributions (\S\ref{sec:synthesis}).
\item We extend the Boomerang implementation with simple symmetric lenses,
  showing that they can easily be combined with a rich variety of other lens
  extensions including quotients and matching lenses. We evaluate our
  implementation over a set of benchmarks drawn from data cleaning, information
  extraction, and file synchronization tasks and demonstrate that we can
  synthesize all of them in under half a minute (\S\ref{sec:evaluation}).
\end{itemize}

@@

In this section, we give an overview of simple symmetric lenses and
our corresponding synthesis algorithm using an example drawn from a
hypothetical U.S. company.  In this company, management and
human resources (HR) store information about employees in separate files:
management records the names of employees and their salaries while
HR stores the names of employees and their health
insurance providers.  Figure~\ref{fig:minimized-representations} gives
examples of the two types of files and regular expressions describing
the two formats.
The company uses simple symmetric lenses to keep these files synchronized.
If management adds a new employee, ``Chris Roe: 32500'',  a lens adds 
the corresponding entry ``Chris Roe, UNK'' to HR's file.  The
sentinel value ``UNK'' represents the fact that the employee's
insurance company is currently unknown.  A similar update happens if
HR adds a new employee before management, in which case the sentinel
value ``N/A'' reflects the unknown salary information. 
Furthermore, if HR
corrects an error in an employee's name, say changing ``John Public''
to ``Jon Public'', the lens mirrors that name change in management's
file. Not all the data is mirrored, however. The management file 
is not updated in response to insurance changes, and the HR file
is oblivious to salary changes.  Simple symmetric lenses are
appropriate for keeping these two files synchronized because the files have
a mix of shared and unshared information.

@@

\subsection{Simple Symmetric Lenses}
A simple symmetric lens between formats described by types $\Regex$
and $\RegexAlt$ 
is defined by four functions

@@

In contrast to the identity lens, the disconnect lens \Disconnect leaves data
unconnected. The disconnect lens takes four arguments: two regular 
expressions and two strings. The regular expressions specify the
formats of the corresponding data on the two sides while the strings provide default
values.  In our extended example, 
the \lstinline{salary} and \lstinline{company} fields are independent,
so we use the disconnect lens to relate them.
The defaults are strings that indicate the information is missing.
%
\begin{lstlisting}
disc(salary, company, "N/A", "UNK") : salary $\Leftrightarrow$ company
\end{lstlisting}
%
The insert \lstinline{ins} and delete \lstinline{del} lenses are
syntactic sugar for uses of the disconnect lens in which a string
constant is omitted entirely from the source and target formats,
respectively. \ksf{Are these in the right order?}
\begin{lstlisting}
 ins($t$) = disc("", $t$, "", $t$)
 del($s$) = disc($s$, "", $s$, "")
\end{lstlisting}

@@

    $\EntropyOf{\PRegexOr{\Regex}{\RegexAlt}{\Probability}}$
    & =
    & $\Probability*(\EntropyOf{\Regex} - \Log_2(\Probability)) + (1-\Probability)*(\EntropyOf{\RegexAlt} - \Log_2(1-\Probability))$\\
  \end{tabular}
\end{center}
We have proven that this syntactic computation corresponds to the entropy of the
stochastic regular expression.

@@

    $\LEntropyOf{\RegexAlt \Given \InvertOf{\Lens}, \Regex}$
    & =
    & $\REntropyOf{\RegexAlt \Given \Lens, \Regex}$\\
  \end{tabular}
\end{center}
$\REntropyOf{\RegexAlt \Given \Lens, \Regex}$ is defined symmetrically.  These
functions calculate the expected number of bits to recover one data format from
a synchronized string in the other format.
The cost of a lens, $c(\Lens \OfType \Regex \Leftrightarrow \RegexAlt) =
\LEntropyOf{\Regex \Given \Lens, \RegexAlt} + \REntropyOf{\RegexAlt \Given
  \Lens, \Regex}$ is the
sum of recovering the left format from the right, and the right from the left.
We have proven theorems demonstrating the calculated entropy corresponds to the
actual conditional entropy to recover the data.

@@

We have a means of turning stochastic regular expressions into stochastic DNF
regular expressions. The conversion algorithm itself, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
We have proven this conversion respects languages and probability distributions.

>>>>

Lens languages, like any domain-specific language, also cut down the space of
possible programs one can write, and the smaller the space of programs,
the easier it is to find the program one is looking for.  
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician, maina+:quotient-synthesis} has exploited this fact by demonstrating that it
is possible to synthesize relatively large, complex lenses from specifications.
In particular, given types that describe source and target formats, there are
synthesis algorithms\bcp{more than one?  I know there are two papers, but
  it's really the same synthesis algorithm} that will find bijective
lenses
% ~\cite{optician}
for files 
in ad-hoc formats, such as those housing linux application configurations.
Oftentimes, these algorithms find such lenses in less than a second, and without
the aid of any examples at all, with just types describing the source and target
formats. 
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}

@@

The second difficulty with synthesizing symmetric lenses is more
fundamental\dpw{better word?}: While the number of bijective lenses between two
related formats is often tiny, the number of symmetric lenses is often large. In
practice, the difference in cardinality means that when a search algorithm
selects the first bijective lens it finds, that lens has a good chance of being
the one a user wants. In contrast, and to our surprise, our experiments show
that if our search algorithm selects the first symmetric lens it finds, it
typically fails to find the lens the user wants. The flexibity these lenses
admit when it comes to eliding information generates an explosion of possible
transformations. To disambiguate between the many possible lenses, we need new
principles for identifying the ``likely'' ones, and we need new algorithms to
search the space of program transformations using these principles. \dpw{Were
  the bijective lenses chosen using a ``smaller'' criteria? I forget... if not
  edit the prior paragraph.}\afm{no, chosen using a ``first'' criteria.
  Edited.}\dpw{Ideally, we could back up the claim that there are many symmetric
  lenses and not many bijective lenses by graphing the number of such lenses
  somehow. eg, in a past synthesis paper with Steve (POPL 2016) we graphed the
  untyped terms of type $\mathsf{int}\rightarrow\mathsf{int}$. I wonder if we
  can do something similar here.}\afm{What about lenses between ``a''* and
  ''b``*}\bcp{Aren't there an infinite number of those, even if we restrict
  to bijective ones?}

To overcome this second difficulty, we turned to information theory to try to
find a way to characterize the likely and unlikely lenses. Intuitively, a
``likely'' lens is one that propagates a lot of information from source to
target and also from target back to source. Conversely, given the source, the
target is ``unlikely'' to require very much additional information in order to
reconstruct it---if it were\bcp{were what? likely, or unlikely?}, then one is
unlikely to be attempting to synthesize a lens between the two formats\bcp{why?
  even if two big formats only share a little bit of common information,
  mightn't one want a lens between them?}! In order to implement this intuition,
we turn to \emph{stochastic regular expressions} (SREs). Stochastic regular
expressions define a set of strings, as do ordinary regular expressions, but
they also provide the expected distribution of strings within that set. Such
distributions allow us to compute the amount of information gained when
presented an element of the set. For example, the language of the SRE
$\lstinline{"a"} |_{0.5}
\lstinline{"b"}$ contains two elements (the strings
\lstinline{"a"} and
\lstinline{"b"}). The probability 0.5 annotating the union indicates that both
elements are equally likely to appear. Consequently, if we observe an element of
that language (either \lstinline{"a"} or \lstinline{"b"}), we will gain one bit
of information. Said another way, one bit, together with the SRE, suffices to
recover the string in question.\bcp{Is this implying that, if the annotation
  were different, we'd need something other than one bit to recover the
  string??}
\afm{From an information theoretic perspective, yes, because that perspective
  views it as the amount of information if we were communicating a stream of
  these choices.}\bcp{I see---very interesting.  What about mentioning this
  point, to get people ready for some information-theoretic thinking
  later...?} 
On the other hand, consider the language of the SRE \lstinline{c}. It contains 
just one element (the string $c$). If we observe that our algorithm generates
$\lstinline{"c"}$, we gain zero bits of information --- the string was already determined.
Hence\bcp{?}, a lens is ``likely,'' in a technical sense, if relative to the
target SRE $T$, very little information from $t$ is required to generate it.

@@

In summary, our main contributions are:
\begin{itemize}
\item We define the space of simple symmetric lenses, which lies between
  asymmetric lenses and full symmetric lenses and has the useful property that
  its members don't have any internal state (\S\ref{sec:overview}). 
\item We show how to synthesize simple symmetric lenses (including all asymmetric lenses as a special case),
using a novel application of stochastic regular expressions to guide the search process.  In particular, we
show that the RE star semi-ring equivalences remain valid when their semantics
are enriched by distributions (\S\ref{sec:synthesis}).
\item We extend the Boomerang implementation with simple symmetric lenses,
  showing that they can easily be combined with a rich variety of other lens
  extensions including quotients and matching lenses. We evaluate our
  implementation over a set of benchmarks drawn from data cleaning, information
  extraction, and file synchronization tasks and demonstrate that we can
  synthesize all of them in under half a minute (\S\ref{sec:evaluation}).
%
  We also analyze the gap between simple and full symmetric lenses
  empirically and argue that simple ones suffice for many applications.
\item We show that the class of simple lenses can be characterized
  semantically as the subset of full symmetric lenses for which a simple
  ``forgetfulness'' property holds (\S\ref{sec:related}).  \bcp{I suggest
    splitting this out into its own section, since there is other related
    work that we'll want to discuss in a Related Work section.}
\end{itemize}

@@

In this section, we give an overview of simple symmetric lenses and
our corresponding synthesis algorithm using an example drawn from a
hypothetical U.S. company.  In this company, management and
human resources (HR) store information about employees in separate files:
management records the names of employees and their salaries while
HR stores the names of employees and their health
insurance providers.  Figure~\ref{fig:minimized-representations} gives
examples of the two types of files and regular expressions describing
the two formats.
The company uses simple symmetric lenses to keep these files synchronized.
If management adds a new employee, ``Chris Roe: 32500'',  a lens adds 
the corresponding entry ``Chris Roe, UNK'' to HR's file.  The
sentinel value ``UNK'' represents the fact that the employee's
insurance company is currently unknown.  A similar update happens if
HR adds a new employee before management, in which case the sentinel
value ``N/A'' reflects the unknown salary information\bcp{can we just use
  ``UNKNOWN'' on both sides?  Or is there some reason to make it more
  complicated?} 
Furthermore, if HR
corrects an error in an employee's name, say changing ``John Public''
to ``Jon Public'', the lens mirrors that name change in management's
file. Not all the data is mirrored, however. The management file 
is not updated in response to insurance changes, and the HR file
is oblivious to salary changes.  Simple symmetric lenses are
appropriate for keeping these two files synchronized because the files have
a mix of shared and unshared information.

@@

\subsection{Simple Symmetric Lenses}
A simple symmetric lens between formats described by types $\Regex$
and $\RegexAlt$ 
is defined by four functions\bcp{I think a picture would be helpful.  I can probably
  make one pretty easily by tweaking the one from the original symmetric paper.}

@@

In contrast to the identity lens, the disconnect lens, \Disconnect, leaves data
unconnected.\bcp{Can we call it ``disconnect''?} The disconnect lens takes four arguments: two regular 
expressions and two strings. The regular expressions specify the
formats of the corresponding data on the two sides while the strings provide default
values.  In our extended example, 
the \lstinline{salary} and \lstinline{company} fields are independent,
so we use the disconnect lens to relate them.
The defaults are strings that indicate the information is missing.
%
\begin{lstlisting}
disc(salary, company, "N/A", "UNK") : salary $\Leftrightarrow$ company
\end{lstlisting}
%
The insert \lstinline{ins} and delete \lstinline{del} lenses are
syntactic sugar for uses of the disconnect lens in which a string
constant is omitted entirely from the source and target formats,
respectively. \ksf{Are these in the right order?}
\begin{lstlisting}
 ins($t$) = disc("", $t$, "", $t$)
 del($s$) = disc($s$, "", $s$, "")
\end{lstlisting}

@@

    $\EntropyOf{\PRegexOr{\Regex}{\RegexAlt}{\Probability}}$
    & =
    & $\Probability*(\EntropyOf{\Regex} - \Log_2(\Probability)) + (1-\Probability)*(\EntropyOf{\RegexAlt} - \Log_2(1-\Probability))$\\
  \end{tabular}
\end{center}
We have proven\bcp{This is a POPL paper: Proofs of important theorems should
be included.} that this syntactic computation corresponds to the entropy of the
stochastic regular expression.

@@

    $\LEntropyOf{\RegexAlt \Given \InvertOf{\Lens}, \Regex}$
    & =
    & $\REntropyOf{\RegexAlt \Given \Lens, \Regex}$\\
  \end{tabular}
\end{center}
$\REntropyOf{\RegexAlt \Given \Lens, \Regex}$ is defined symmetrically.  These
functions calculate the expected number of bits to recover one data format from
a synchronized string in the other format.
The cost of a lens, $c(\Lens \OfType \Regex \Leftrightarrow \RegexAlt) =
\LEntropyOf{\Regex \Given \Lens, \RegexAlt} + \REntropyOf{\RegexAlt \Given
  \Lens, \Regex}$ is the
sum of recovering the left format from the right, and the right from the left.
We have proven theorems demonstrating the calculated entropy corresponds to the
actual conditional entropy to recover the data.\bcp{Again, they should be
  included.}

@@

We have a means of turning stochastic regular expressions into stochastic DNF
regular expressions. The conversion algorithm itself, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
We have proven this conversion respects languages and probability
distributions.\bcp{Should be included.}
